\bf An observational study of papers published in \emph{PLOS ONE} and studies posted to a trial registry.

### 2021-05-24 manuscript/asa_template/asa_template_almostdone.tex
add edited tex file

The statistical methods section for the ANZCTR data was missing for some
studies and we examined if there were particular studies where this
section was more likely to be missing. We used four independent
variables of date, study type (observational or interventional), number
of funders and target sample size. We used a logistic regression model
fitted using a Bayesian paradigm. A small number of sections were
labelled as ``Not applicable'', ``Nil'' or ``None'' and we changed these
to missing.

@@

To address these limitations, NMF deals with high-dimensional data by
mapping it to a lower-dimensional space. This mapping this achieved by
approximating \(P\) with two factor matrices: \(W \in R^{M \times g}\)
and \(H \in R ^{N \times g}\) \citep{aggarwal2012mining}. The number of
subgroups of common topic inferred from the data is represented by
\(g\), leading to the approximation: \begin{equation}
\label{eq:approxPNMF}
 P \approx WH^{T}.
\end{equation}

@@

Following estimation, \textit{H} contains the information regarding
topic membership for all studies. In our case, topic membership
\((1,\ldots,g)\) for a statistical methods section is inferred from the
maximum coefficient value in the corresponding row of \(H\).

For our two datasets, we applied NMF varying \(g\) from 1 to 50.
Clustering solutions were assessed using silhouette score,
within-cluster dispersion and between-cluster dispersion.

\subsection{Within topic analysis}

Within the \emph{PLOS ONE} dataset, statistical methods sections with
high topic coherence scores were used to investigate evidence of
boilerplate text, compared with other studies assigned to the same
topic. Comparisons were carried out by calculating Jaccard similarity
scores between studies at the sentence level. We chose the Jaccard score
as an intuitive measure, which summarises the similarity between two
sentences by the number of words common to each sentence (intersection),
divided by the number of words that appeared in either the strongest
match or the sentence being compared. Boilerplate text was defined based
on word count (plus or minus 3 words) and a Jaccard score of 0.9 or
higher.

Results were transformed to lower case for the clustering, but examples
are given using the original capitalisation.

@@

\hypertarget{included-studies}{%
\subsection{Included studies}\label{included-studies}}

\hypertarget{plos-one}{%
\subsubsection{PLOS ONE}\label{plos-one}}

@@

\caption{\label{fig:plos-n}PLOS summary}\label{fig:unnamed-chunk-3}
\end{figure}

Searching with the \emph{PLOS ONE} API returned 131,847 results (DOIs)
based on chosen search terms (\autoref{fig:plos-flowchart}). After
partial matching, 111,731 (85\%) statistical methods sections were
identified. In the final sample, 95,518 (85\%) DOIs returned an exact
match against common section headings: 64,133 for `statistical
analysis', 13,380 for `statistical analyses' and 13,627 for `data
analysis'. For our random sample of DOIs that did not meet the partial
matching criteria, initial search terms appeared in Introduction
(number), Results (number) only. xxx appeared in the Methods section
without a section heading and xxx had a non-standard section heading.
{[}TODO{]}.

Search results varied by journal volume (Figure \ref{fig:plos-n}A). The
total number of API search results peaked at volumes 8 (n = 19,045) and
9 (n = 19,045) in years 2013 and 2014. This trend aligned with the total
number of papers published in \emph{PLOS ONE} over the same period. The
percentage of records that included a statistical methods section by
volume based on our proposed matching criteria varied between 64\%
(volume 2) and 86\% (volume 9).

All DOIs included Biology and life sciences (n = 107,584), Earth
sciences (n = 7,605) and/or Computer and information sciences (n =
5,190) in their top 3 subject classifications (Figure
\ref{fig:plos-n}B).

@@

@@

\caption{\label{fig:plos-clouds}Topic clouds for ten clusters for PLOS ONE}\label{fig:unnamed-chunk-5}
\end{figure}

The topic clouds based on ten clusters are in Figure
\ref{fig:plos-clouds}. Frequently occurring words reflected the use of
statistical software (topics 3 and 5), descriptive statistics (topic 6),
group based hypothesis testing (topics 1 and 4) and definitions of
statistical significance (topics 1 and 9). Statistical methods sections
related to regression (topic 2) and meta-analysis (topic 7) were also
identified.

Topics related to statistical software differentiated between Prism
GraphPad (topic 3: n = 9974; 8.6\%) and SPSS (topic 5: n = 9648; 8.3\%)
. Manual review of results showed that nine out of the ten top matching
DOIs for topic 3 stated the use of Prism GraphPad, but did not specify
which statistical methods were used. Top matching sections for topic 5
included information on SPSS version numbers and definitions of
statistical significance.

\begin{table}

\caption{\label{tab:unnamed-chunk-7}Example boilerplate text from PLOS ONE dataset}
\centering
\begin{tabular}[t]{l|l|r}
\hline
Topic & Statistical methods text & Number of matching studies\\
\hline
Topic 1 & a p value of less-than 0.05 was considered statistically significant. & 1883\\
\hline
Topic 1 & p values of less-than 0.05 were considered significant. & 1526\\
\hline
Topic 1 & statistical analysis was performed using the students t-test & 380\\
\hline
Topic 1 & students t-test was used for statistical analysis & 29\\
\hline
Topic 3 & graphpad prism graphpad software san diego ca was used for all analyses. & 1941\\
\hline
Topic 3 & all statistical analysis was performed using graphpad prism software. & 1312\\
\hline
Topic 4 & statistical analysis was performed using one-way analysis of variance anova test followed by dunnetts multiple comparison post hoc test & 4516\\
\hline
Topic 4 & p values less-than 0.05 were considered significant. & 2414\\
\hline
Topic 4 & significant differences were determined using analysis of variance anova followed by tukey post-hoc tests for multiple comparisons. & 2216\\
\hline
Topic 4 & data are expressed as the mean plus-or-minus sem & 860\\
\hline
Topic 5 & the difference was considered statistically significant at p less-than 0.05. & 2555\\
\hline
Topic 5 & spss software version 17.0 spss chicago il usa was used for statistical analysis. & 441\\
\hline
Topic 5 & the results are presented as the mean plus-or-minus sem & 420\\
\hline
Topic 5 & data were analyzed by one-way anova and lsd tests using the spss 13.0 software spss inc chicago il usa & 19\\
\hline
Topic 6 & a value of p less-than 0.05 was considered statistically significant. & 1786\\
\hline
Topic 6 & data are expressed as mean plus-or-minus standard error of the mean sem & 604\\
\hline
Topic 6 & statistical analysis of the data was performed by the students t-test & 444\\
\hline
Topic 6 & all results are expressed as means plus-or-minus standard deviation plus-or-minus sd. & 102\\
\hline
Topic 9 & statistical analysis was performed using the students unpaired t-test & 752\\
\hline
Topic 9 & differences were considered statistically significant at p less-than 0.05 p less-than 0.01 and p less-than 0.001. & 413\\
\hline
Topic 9 & data are presented as the mean plus-or-minus sd & 378\\
\hline
Topic 9 & statistical significance was determined by students t-tests & 49\\
\hline
Topic 9 & p less-than 0.05 p equal-to less-than 0.01 p less-than 0.001. & 6\\
\hline
\end{tabular}
\end{table}

Definitions of statistical significance featured strongly in topic 1 (n
= 3784; 3.2\%) and topic 9 (n = 6195; 5.3\%), combined with text
describing hypothesis testing for comparing differences between two
groups. Topic 1 reflected applications of Student's t-test assuming a
5\% level of statistical significance. Topic 9 referenced similar
methods combined with multiple thresholds for declaring statistical
significance by asterisk: ``\(^{*}p<0.05\), \(^{**}p<0.01\) and
\(^{***}p<0.001\)'', a practice that has been criticised
\citep{Wasserstein2019}.

Group-based hypothesis testing was a recurring theme across topics, with
text descriptions varying based on method(s) used. One-way analysis of
variance also featured strongly in topic 4 (n = 10212; 8.8\%), combined
with common methods for performing post-hoc multiple comparisons.
Frequently occurring words in topic 6 (n = 4764; 4.1\%) reflected
mentions of descriptive statistics for summarising continuous variables,
for example: ``Data are expressed as means ± standard error of the mean
(SEM).'' Sections assigned to this topic appeared to be expanded
versions of topics 1 and 9. Examples of boilerplate text were in the
form of descriptive statistics followed by hypothesis testing, for
example, Student's t-test, Mann Whitney U or one-way analysis of
variance.

Examples of boilerplate text for selected topics are presented in Table
\ref{tab:plos-example-boilerplate}.

@@

We downloaded 28,008 studies. The numbers of excluded studies are shown
in Figure \ref{fig:anzctr-missing}. Of the 12,700 included studies,
9,523 (75\%) had a statistical methods section. The median length of
sections was 129 words with an inter-quartile range of 71 to 219 words.

We examined if four study characteristics were associated with a missing
statistics section. The odds ratios and 95\% credible intervals are in
Table \ref{tab:anzctr-missing-odds}. Observational studies were less
likely to have a missing methods section compared with interventional
studies. Missing sections became less likely over time. Studies with
more funders and a larger target sample size were less likely to have a
missing methods section.

\begin{table}[]
\centering
\caption{Regression results for factors associated with missing statistical methods sections in ANZCTR}
\label{tab:anzctr-missing-odds}
\begin{tabular}{lcc}
\hline
Variable & Odds ratio & 95$\%$ CI \\
\hline
Study type $=$ Observational & 0.78 & (0.69, 0.89) \\
Date (per year) & 0.90 & (0.88, 0.91) \\
Number of funders & 0.80 & (0.74, 0.86) \\
Target sample size (per doubling) & 0.90 & (0.88, 0.92)\\
\hline
\end{tabular}
\end{table}

{\centering \includegraphics[width=0.7\linewidth]{figures/excluded_anzctr_missing}

\caption{\label{fig:anzctr-missing}ANZCTR search results.}\label{fig:unnamed-chunk-8}
\end{figure}

@@

The clustering algorithm found groups that were purely sample size
calculations (topic 2, n = 1834; topic 4, n = 909), pilot studies (topic
5, n = 834), safety/tolerability studies (topic 6, n = 524),
intervention studies (topic 8, n = 1020) and repeated measures ANOVA
(topic 10 = 852). Examples of boilerplate text are provided in Table
\ref{tab:anzctr-example-boilerplate} and Supplementary File 3.

@@

Finally, targeted searches for the term `statistician' revealed
instances across all topics where no methods were specified. Instead,
all analyses were the responsibility of an appointed statistician, for
example, ``A statistician employed by hospital was used'' and ``Pilot
study at this point will use a statistician professionally to determine
sample size calculations as required''.

>>>>

Statistical methods sections were missing for some studies downloaded
from ANZCTR, including a small number of sections labelled as ``Not
applicable'', ``Nil'' or ``None''. We therefore examined if there were
particular studies where this section was more likely to be missing.
Analysis considered a logistic regression model estimated in the
Bayesian framework {[}\citet{INLA}; www.r-inla.org{]}, with missing
statistical methods section (yes/no) as the dependent variable. Date,
study type (observational or interventional), number of funders and
target sample size were used as independent variables. Results were
reported as odds ratios with 95\% credible intervals (CI).

@@

To address these limitations, NMF deals with high-dimensional data by
mapping it to a lower-dimensional space. This mapping this achieved by
approximating \(P\) with two factor matrices: \(W \in R^{M \times g}\)
and \(H \in R ^{N \times g}\) \citep{aggarwal2012mining}, such that
\(P \approx WH^{T}\). The number of subgroups of common topic inferred
from the data is given by \(g\).

@@

Following estimation, \textit{H} contains the information regarding
topic membership for all studies. In our case, topic membership
\((1,\ldots,g)\) for a statistical methods section is inferred from the
maximum coefficient value in the corresponding row of \(H\), also known
as the topic coherence score. For our two datasets, we applied NMF with
\(g=10\).

\subsection{Content analysis}

Clustering results were visualised by word clouds to summarise
frequently occurring terms associated with topic membership. Whilst
these results were useful for inferring common themes across statistical
methods sections, they did not indicate evidence of boilerplate text
beyond the use of common terms. Follow up analysis therefore considered
similarities in text between sections assigned to the same topic.

We used the Jaccard similarity index to compare the statistical methods
section with the highest topic coherence score with all other sections
assigned to the same topic. Indices were calculated at the sentence
level, to account for instances of boilerplate text within larger
methods sections. We chose the Jaccard index as an easy to interpret
measure, which summarises the similarity between two sentences by the
number of words common to each sentence (intersection), divided by the
number of words that appeared in either sentence (union). Instances of
boilerplate text were defined by a Jaccard index of 0.9 or higher and a
difference in word count of plus or minus three words.

@@

\hypertarget{plos-one}{%
\subsection{PLOS ONE}\label{plos-one}}

{\centering \includegraphics[width=0.49\linewidth]{figures/excluded_plosone} \includegraphics[width=0.49\linewidth]{figures/excluded_anzctr_missing}

\caption{\label{fig:consort-diagrams}Search results for PLOS ONE (left) and ANZCTR (right).}
\end{figure}

@@

\caption{\label{fig:plos-n} A: Search results by PLOS ONE volume; B: Subject classifications assigned to full-text recorded included in the analysis}
\end{figure}

API searches returned 131,847 results (DOIs)
(\autoref{fig:consort-diagrams}). After partial matching, 111,731 (85\%)
statistical methods sections were identified. In the final sample,
95,518 (85\%) DOIs returned an exact match against common section
headings: 64,133 for `statistical analysis', 13,380 for `statistical
analyses' and 13,627 for `data analysis'.

Search results varied by journal volume (Figure \ref{fig:plos-n}A). The
total number of API search results peaked at volumes 8 (n = 19,045) and
9 (n = 19,045) in years 2013 and 2014. This trend aligned with the total
number of papers published in \emph{PLOS ONE} over the same period. The
percentage of records that included a statistical methods section by
volume based on our proposed matching criteria varied between 64\%
(volume 2) and 86\% (volume 9). All DOIs included Biology and life
sciences (n = 107,584), Earth sciences (n = 7,605) and/or Computer and
information sciences (n = 5,190) in their top 3 subject classifications
(Figure \ref{fig:plos-n}B).

@@

For studies excluded based on section headings, initial search terms
appeared in Introduction (number), Results (number) only. xxx appeared
in the Methods section without a section heading and xxx had a
non-standard section heading. {[}TODO{]}.

@@

\caption{\label{fig:plos-clouds}Topic clouds for ten clusters for PLOS ONE}\label{fig:unnamed-chunk-6}
\end{figure}

The topic clouds for ten clusters are in Figure \ref{fig:plos-clouds}.
Frequently occurring words reflected the use of statistical software
(topics 3 and 5), descriptive statistics (topic 6), group based
hypothesis testing (topics 1 and 4) and definitions of statistical
significance (topics 1 and 9). Statistical methods sections related to
regression (topic 2) and meta-analysis (topic 7) were also identified.
Examples of boilerplate text for selected topics are presented in Table
\ref{tab:plos-example-boilerplate}.

Topics related to statistical software differentiated between Prism
GraphPad (topic 3: n = 9974; 8.6\%) and SPSS (topic 5: n = 9648; 8.3\%).
Within topic 3, the sentence ``GraphPad Prism (Graphpad Software, San
Diego, CA) was used for all analyses'' was found in 1,941 studies.
Similar phrasing was identified for topic 5: ``SPSS software version
17.0 (SPSS, Chicago, IL, USA) was used for statistical analysis'', with
similar text found in 441 studies.

\begin{landscape}
\begin{table}

\caption{\label{tab:plos-example-boilerplate}Example boilerplate text from PLOS ONE datase (sentence level). Statistical methods text corresponds to the DOI with the highest topic coherence score within a topic}
\centering
\begin{tabular}[t]{p{0.1\textwidth}p{0.9\textwidth}p{0.1\textwidth}}
\hline
Topic & Statistical methods text & Matching DOIs\\
\hline
Topic 1 & Students t-test was used for statistical analysis. & 29\\
\cline{2-3}
 & A p value of $<$ 0.05 was considered statistically significant. & 1883\\
\hline
Topic 3 & GraphPad Prism (Graphpad Software, San Diego, CA) was used for all analyses. & 1941\\
\hline
Topic 4 & Significant differences were determined using analysis of variance (ANOVA) followed by tukey post-hoc tests for multiple comparisons. & 2216\\
\hline
Topic 5 & SPSS software version 17.0 (SPSS, Chicago, IL, USA) was used for statistical analysis. & 441\\
\hline
Topic 6 & All results are expressed as means $\pm$ standard deviation ($\pm$ SD). & 102\\
\hline
Topic 9 & Statistical significance was determined by Students t-tests. & 49\\
\cline{2-3}
& $^*$p $<$ 0.05, $^{**}$p $<$ 0.01, $^{***}$p $<$ 0.001. & 6\\
\hline
\end{tabular}
\end{table}
\end{landscape}

Definitions of statistical significance featured strongly in topic 1 (n
= 3784; 3.2\%) and topic 9 (n = 6195; 5.3\%). Topic 1 reflected
applications of two-tailed and unpaired Student's t-tests, however,
instances of boilerplate text emphasised the 5\% level of statistical
significance (\ref{tab:plos-example-boilerplate}). In contrast, Topic 9
focused on multiple thresholds for declaring statistical significance by
asterisk: ``\(^{*}p<0.05\), \(^{**}p<0.01\) and \(^{***}p<0.001\)'', a
practice that has been criticised \citep{Wasserstein2019}.

Group-based hypothesis testing was a recurring theme across topics, with
text descriptions varying based on method(s) used. One-way analysis of
variance also featured strongly in topic 4 (n = 10212; 8.8\%), combined
with common methods for performing post-hoc multiple comparisons. Based
on the Jaccard index, 1 in 5 studies were matched to the sentence
``Significant differences were determined using analysis of variance
(ANOVA) follwo by Tukey post-hoc tests for multiple comparisons''.
Frequently occurring words in topic 6 (n = 4764; 4.1\%) reflected
mentions of descriptive statistics for summarising continuous variables,
namely means with standard deviations or standard errors.

@@

We downloaded 28,008 studies. The numbers of excluded studies are shown
in Figure \ref{fig:consort-diagrams}. Of the 12,700 included studies,
9,523 (75\%) had a statistical methods section. The median length of
sections was 129 words with an inter-quartile range of 71 to 219 words.

Odds ratios and 95\% credible intervals for study characteristics
included in the logistic regression model are presented in Table
\ref{tab:anzctr-missing-odds}. Observational studies were less likely to
have a missing statistical methods section compared with interventional
studies. Missing sections became less likely over time. Studies with
more funders and a larger target sample size were less likely to have a
missing statistical methods section.

\begin{table}[]
\centering
\caption{Logistic regression results for study characteristics associated with missing statistical methods sections in ANZCTR}
\label{tab:anzctr-missing-odds}
\begin{tabular}{lcc}
\hline
Variable & Odds ratio & 95$\%$ CI \\
\hline
Study type $=$ Observational & 0.78 & (0.69, 0.89) \\
Date (per year) & 0.90 & (0.88, 0.91) \\
Number of funders & 0.80 & (0.74, 0.86) \\
Target sample size (per doubling) & 0.90 & (0.88, 0.92)\\
\hline
\end{tabular}
\end{table}

@@

The clustering algorithm found groups that were purely sample size
calculations (topic 2, n = 1834), pilot studies (topic 5, n = 834),
safety/tolerability studies (topic 6, n = 524), intervention studies
(topic 8, n = 1020) and repeated measures ANOVA (topic 10 = 852).
Examples of boilerplate text are provided in Table
\ref{tab:anzctr-example-boilerplate} .

@@

Since studies included in this dataset outlined planned analyses, we
suspected that some studies did not specify statistical methods because
they had yet to consult with a statistician. Targeted searches for the
term ``statistician'' across all topics returned xx matching records,
with examples including ``A statistician employed by hospital was used''
and ``Pilot study at this point will use a statistician professionally
to determine sample size calculations as required''.
